---
title: 'Blog Post 3: Polling'
author: Alex Heuss
date: '2024-09-23'
slug: blog-3-polling
categories: []
tags: []
---
```{r, load packages, echo=FALSE, message=FALSE}
library(car)
library(caret)
library(maps)
library(CVXR)
library(glmnet)
library(tidyverse)
```

```{r, load data, echo=FALSE, message=FALSE}
nat_poll_data <- read_csv("data/national_polls_1968-2024.csv")
state_av_poll_data <- read_csv("data/state_polls_1968-2024.csv")
national_pop_vote_data <- read_csv("data/popvote_1948-2020.csv") 
nat_pop_party_neutral <- 
  national_pop_vote_data |>
  mutate(dem_pv = ifelse(party == "democrat", pv, 0),
         rep_pv = ifelse(party == "republican", pv, 0),
         dem_2pv = ifelse(party == "democrat", pv2p, 0),
         rep_2pv = ifelse(party == "republican", pv2p, 0)) |>
  group_by(year) |>
  summarize(dem_pv = sum(dem_pv),
            rep_pv = sum(rep_pv),
            dem_2pv = sum(dem_2pv),
            rep_2pv = sum(rep_2pv))

```

## Historic Accuracy of Polls

Poll averages seem like an intuitive way to predict elections - intuitively, if I walk around and ask everyone who they are going to vote for, I should get a pretty good idea of how many people are voting for each candidate. However, historically, polls become more accurate as the election approaches, often converging near the correct value within a week or so of the election. Several theories attempt to explain this, including the "enlightened blah blah" theory expressed by so and so in their book *book*.

[insert a graph of this if possible (picturing a scatterplot with weeks_left on the bottom and difference between poll and actual on the y)]



```{r, lots of data wrangling, echo=FALSE, message=FALSE}

nat_poll_data_party_neutral <- 
  nat_poll_data |>
  mutate(dem_poll = ifelse(party=="DEM", poll_support, 0),
         rep_poll = ifelse(party=="REP", poll_support, 0)) |>
  select(-party, -candidate, -poll_support) |>
  group_by(year, weeks_left, poll_date) |>
  summarize(dem_poll = sum(dem_poll),
            rep_poll = sum(rep_poll))

clean_nat_data <- 
  nat_pop_party_neutral |>
  filter(year >= 1968) |>
  left_join(nat_poll_data_party_neutral, by = "year")

weekly_nat_data <- clean_nat_data |>
  group_by(year, weeks_left) |>
  summarize(av_d_pv = mean(dem_pv),
            av_r_pv = mean(rep_pv),
            av_d_2pv = mean(dem_2pv),
            av_r_2pv = mean(rep_2pv),
            av_d_poll = mean(dem_poll, na.rm = TRUE),
            av_r_poll = mean(rep_poll, na.rm = TRUE))

poll_data_for_merge <- nat_poll_data |> 
  mutate(party = ifelse(party == "DEM", "democrat", "republican")) |>
  group_by(year, party) |> 
  top_n(1, poll_date) |> 
  select(-candidate)

nov_data <- 
  national_pop_vote_data |> 
  left_join(poll_data_for_merge,
            by = c("year", "party")) |>
  rename(nov_poll = poll_support) |>
  filter(year <= 2020) |>
  drop_na()
```

```{r, graph last poll preds and the actual result, echo=FALSE}
# literally cannot figure out why this isn't working RIP
nov_data |>
  mutate(Party = ifelse(party == "democrat", "Democrat", "Republican")) |>
  ggplot(aes(x = nov_poll, y = pv, color = Party)) +
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 0.75, se=FALSE) +
  geom_abline(slope = 1) + 
  geom_hline(yintercept = 50) +
  geom_vline(xintercept = 50) +
  scale_color_manual(values = c("#00AEF3", "#E81B23")) +
  labs(x = "Latest Poll Result (%)", 
       y = "National Popular Vote Share (%)", 
       title = "Final Poll Predictions and National Popular Vote Share (1968-2020)") +
  theme_bw()


```

## Visualize 2024 Polls

```{r, 2024 nat polls graph, echo=FALSE}
nat_poll_data |> 
  filter(year == 2024 & (poll_date > '2024-07-21' | poll_date == '2024-07-21')) |> 
  mutate(Candidate = ifelse(party == "DEM", "Kamala Harris", "Donald Trump")) |>
  ggplot(aes(x = poll_date, y = poll_support, color = Candidate)) +
  geom_point(size = 1) + 
  geom_line() + 
  scale_x_date(date_labels = "%b %d") + 
  scale_color_manual(values = c("#E81B23", "#00AEF3")) +
  labs(x = "Date",
       y = "Average Poll Approval", 
       title = "Polling Averages Post-Biden Dropout") + 
  theme_bw()

```
## State Level
```{r, load in more data and merge, echo=FALSE, message=FALSE}
state_pop_data <- read_csv("data/clean_wide_state_2pv_1948_2020.csv")
ec <- read_csv("data/corrected_ec_1948_2024.csv")
```


```{r. state data wrangling, echo=FALSE, message=FALSE}
clean_state_polls <- 
  state_av_poll_data |>
  mutate(party = ifelse(party == "DEM", "democrat", "republican"),
         dem_support = ifelse(party == "democrat", poll_support, 0),
         rep_support = ifelse(party == "republican", poll_support, 0)) |>
  select(-party, -candidate, -days_left, -poll_support) |>
  group_by(year, weeks_left, state, poll_date) |>
  summarize(dem_support = sum(dem_support),
            rep_support = sum(rep_support))

merged_state_data <- 
  state_pop_data |>
  filter(year >= 1972) |>
  left_join(clean_state_polls, by = c("state", "year")) |>
  select(year, state, D_pv, R_pv, D_pv2p, R_pv2p, weeks_left, poll_date, dem_support, rep_support) |>
  left_join(ec, by = c("state", "year"))

average_state_data <- 
  merged_state_data |>
  group_by(year, state) |>
  summarize(av_d_pv = mean(D_pv),
            av_r_pv = mean(R_pv),
            av_d_2pv = mean(D_pv2p),
            av_r_2pv = mean(R_pv2p),
            av_d_poll = mean(dem_support, na.rm = TRUE),
            av_r_poll = mean(rep_support, na.rm = TRUE))
  
weekly_state_data <- 
  merged_state_data |>
  group_by(year, state, weeks_left) |>
  summarize(av_d_pv = mean(D_pv),
            av_r_pv = mean(R_pv),
            av_d_2pv = mean(D_pv2p),
            av_r_2pv = mean(R_pv2p),
            av_d_poll = mean(dem_support, na.rm = TRUE),
            av_r_poll = mean(rep_support, na.rm = TRUE))

```







