---
title: 'Blog Post 3: Polling'
author: Alex Heuss
date: '2024-09-23'
slug: blog-post-3-polling
categories: []
tags: []
---
```{r, load packages, echo=FALSE, message=FALSE}
library(car)
library(caret)
library(maps)
library(CVXR)
library(glmnet)
library(tidyverse)
```

```{r, load data, echo=FALSE, message=FALSE}
nat_poll_data <- read_csv("data/national_polls_1968-2024.csv")
state_av_poll_data <- read_csv("data/state_polls_1968-2024.csv")
national_pop_vote_data <- read_csv("data/popvote_1948-2020.csv") 
nat_pop_party_neutral <- 
  national_pop_vote_data |>
  mutate(dem_pv = ifelse(party == "democrat", pv, 0),
         rep_pv = ifelse(party == "republican", pv, 0),
         dem_2pv = ifelse(party == "democrat", pv2p, 0),
         rep_2pv = ifelse(party == "republican", pv2p, 0)) |>
  group_by(year) |>
  summarize(dem_pv = sum(dem_pv),
            rep_pv = sum(rep_pv),
            dem_2pv = sum(dem_2pv),
            rep_2pv = sum(rep_2pv))

```

## Historic Accuracy of Polls

Poll averages seem like an intuitive way to predict elections - intuitively, if I walk around and ask everyone who they are going to vote for, I should get a pretty good idea of how many people are voting for each candidate. However, historically, polls become more accurate as the election approaches, often converging near the correct value within a week or so of the election. Several theories attempt to explain this, including the "enlightened blah blah" theory expressed by so and so in their book *book*.

[insert a graph of this if possible (picturing a scatterplot with weeks_left on the bottom and difference between poll and actual on the y)]



```{r, lots of data wrangling, echo=FALSE, message=FALSE}

nat_poll_data_party_neutral <- 
  nat_poll_data |>
  mutate(dem_poll = ifelse(party=="DEM", poll_support, 0),
         rep_poll = ifelse(party=="REP", poll_support, 0)) |>
  select(-party, -candidate, -poll_support) |>
  group_by(year, weeks_left, poll_date) |>
  summarize(dem_poll = sum(dem_poll),
            rep_poll = sum(rep_poll))

clean_nat_data <- 
  nat_pop_party_neutral |>
  filter(year >= 1968) |>
  left_join(nat_poll_data_party_neutral, by = "year")

weekly_nat_data <- clean_nat_data |>
  group_by(year, weeks_left) |>
  summarize(av_d_pv = mean(dem_pv),
            av_r_pv = mean(rep_pv),
            av_d_2pv = mean(dem_2pv),
            av_r_2pv = mean(rep_2pv),
            av_d_poll = mean(dem_poll, na.rm = TRUE),
            av_r_poll = mean(rep_poll, na.rm = TRUE))

poll_data_for_merge <- nat_poll_data |> 
  mutate(party = ifelse(party == "DEM", "democrat", "republican")) |>
  group_by(year, party) |> 
  top_n(1, poll_date) |> 
  select(-candidate)

nov_data <- 
  national_pop_vote_data |> 
  left_join(poll_data_for_merge,
            by = c("year", "party")) |>
  rename(nov_poll = poll_support) |>
  filter(year <= 2020) |>
  drop_na()
```

```{r, graph last poll preds and the actual result, echo=FALSE}

nov_data |>
  mutate(Party = ifelse(party == "democrat", "Democrat", "Republican")) |>
  ggplot(aes(x = nov_poll, y = pv, color = Party)) +
  geom_point() + 
  geom_smooth(method = "lm", linewidth = 0.75, se=FALSE) +
  geom_abline(slope = 1) + 
  geom_hline(yintercept = 50) +
  geom_vline(xintercept = 50) +
  scale_color_manual(values = c("#00AEF3", "#E81B23")) +
  labs(x = "Latest Poll Result (%)", 
       y = "National Popular Vote Share (%)", 
       title = "Final Poll Predictions and National Popular Vote Share (1968-2020)") +
  theme_bw()


```