---
title: Election Forecast 1
author: Alex Heuss
date: '2024-09-05'
slug: testing-blog-upload
categories: [election-prediction]
tags: [president]
---
```{r, setup, echo = FALSE, message = FALSE}
library(ggplot2)
library(tidyverse)
library(maps)

#note: code to build site - build_site(build_rmd=TRUE)
```

## General Exploration

This is my first blog post, for the week of September 2nd. The primary method of prediction for this week is a model based on the results of the past two presidential elections, 2016 and 2020. 

```{r, loading data, echo= FALSE, message = FALSE} 
# The code for this post is heavily based on the code that was introduced in section this week. 

nat_data <- read_csv("popvote_1948-2020.csv")
state_data <- read_csv("clean_wide_state_2pv_1948_2020.csv")

```
```{r, initial data wrangling, echo = FALSE}
nat_wide <- 
  nat_data |> 
  select(year, party, pv2p) |>
  pivot_wider(names_from = party, values_from = pv2p) |>
  mutate(winner = ifelse(republican > democrat, "republican", "democrat"))
  
```

```{r, echo = FALSE}
nat_summary <- 
  nat_wide |> 
  mutate(winner = ifelse(winner == "democrat", "Democrat", "Republican")) |>
  group_by(winner) |>
  summarize(races_won = n())

nat_summary |>
  knitr::kable(col.names = c("Party", "Races Won"), caption = "Presidential races won by each party since 1948")
```

As shown in the table, Democratic presidential candidates have won the popular vote in 11 of the past 19 elections, and Republican candidates 8. Another takeaway from this exploratory table is how little data there actually is to train models on. There have only been 19 presidential elections since 1948, which means that the impacts of outliers in the data may be extra large. 

The graph below maps the two-party vote share for the Democratic and Republic parties since 1948. 

```{r, bar chart, echo= FALSE}

my_theme <- theme_bw() + theme(legend.position = "none")
dem_rep <- c("#00AEF3", "#E81B23")
dem_blue <- "#00AEF3"
rep_red <- "#E81B23"

options(repr.plot.width=20, repr.plot.height=8)

nat_data |>
  ggplot(aes(x = year, y = pv2p, color = party)) +
  geom_line() +
  scale_color_manual(values = dem_rep) + 
  labs(x = "", y = "% Two Party Vote Share", title = "Presidential Two Party Vote Share") + 
  my_theme

```

## 2020 State Maps and Predictions Based on Them

```{r}
states_map <- map_data("state")
state_data <- 
  state_data |>
  mutate(region = tolower(state))

pv_map <- 
  state_data |>
  filter(year == 2020) |>
  left_join(states_map, by = "region")

pv_map |> 
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = R_pv2p), color = "white")+
  scale_fill_gradient(low = "white", high = rep_red) +
  theme_void()

```
```{r}
pv_map_winner <- 
  pv_map |>
  mutate(winner = ifelse(R_pv > D_pv, "republican", "democrat"))

pv_map_winner |>  
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = winner), color = "white") +
  scale_fill_manual(values=dem_rep) +
  theme_void()
```
```{r}
pv_map_margin <- 
  pv_map_winner |>
  mutate(margin = R_pv2p-D_pv2p)

pv_map_margin |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = margin), color = "black") +
  scale_fill_gradient2(high = rep_red,
                       mid = "white",
                       name = "win margin",
                       low = dem_blue,
                       breaks = c(-50,-25,0,25,50),limits=c(-50,50)) +
  theme_void()

```
```{r}
state_2p <- 
  state_data |> 
  filter(year == 2020) |>
  group_by(state) |> 
  summarise(D_2024_2p = 0.75*D_pv2p + 0.25*D_pv2p_lag1,
            R_2024_2p = 0.75*R_pv2p + 0.25*R_pv2p_lag1) |>
  mutate(margin_2024 = R_2024_2p - D_2024_2p,
         region = tolower(state))

state_2p |>
  left_join(states_map, by = "region") |>
  ggplot(aes(long, lat, group = group)) +
  geom_polygon(aes(fill = margin_2024), color = "black") +
  ggtitle("2024 Presidential Forecast (Simplified Electoral Cycle Model)") + 
  scale_fill_gradient2(high = rep_red, 
                       mid = "white",
                       name = "win margin",
                       low = dem_blue, 
                       breaks = c(-50,-25,0,25,50), 
                       limits=c(-50,50)) + 
  theme_void()


```












